Exploring mathematical objects from custom-tailored mathematical universes

### 2020-06-21 paper-filmat.tex
to do
typo (in response to review)

The only caveat of Theorem~\ref{thm:reasoning} is that toposes generally only
support intuitionistic reasoning, not the full power of the ordinary
\emph{classical reasoning}. That is, within most toposes, the law of excluded
middle ($\varphi \vee \neg\varphi$) and the law of double negation elimination
($\neg\neg\varphi \Rightarrow \varphi$) are not available. It is intuitionistic
logic and not classical logic which is the common denominator of all toposes;
we cannot generally argue by contradiction in a topos.

@@

\begin{enumerate}
\item The \emph{trivial topos}. In the trivial topos, any statement whatsoever
is true. The trivial topos is not interesting on its own, but its existence
streamlines the theory and it can be an interesting question whether a given
topos coincides with the trivial topos.
\item $\Set$, the \emph{standard topos}. A statement is true in~$\Set$ iff it is true
in the ordinary mathematical sense.
\item $\Set_M$, the topos associated to any model~$(M,\in)$ of~\textsc{zf}.
\item $\Set^W$, the category of functors~$(W,\leq) \to \Set$ associated to
any Kripke model~$(W,\leq)$. A statement is true in this topos iff it is valid
with respect to the ordinary Kripke semantics of~$(W,\leq)$. This example shows
that the Kripke--Joyal semantics of toposes generalizes the more familiar
Kripke semantics.
\item $\Eff$, the \emph{\effective topos}. A statement is true in~$\Eff$ iff
if it has a \emph{computable witness} as detailed in
Section~\ref{sect:effective-topos}. In~$\Eff$, any function~$\NN \to \NN$ is
computable, any function~$\RR \to \RR$ is continuous and the countable axiom of
choice holds (even if it does not on the meta level).
\item $\Sh(X)$, the \emph{topos of sheaves} over any space~$X$. A statement is true
in~$\Sh(X)$ iff it holds \emph{locally on~$X$}, as detailed in
Section~\ref{sect:sheaf-toposes}. For most choices of~$X$, the
axiom of choice and the intermediate value theorem fail in~$\Sh(X)$, and this
failure is for geometric reasons.
\item $\Zar(A)$, the \emph{Zariski topos} of a ring~$A$ presented in
Section~\ref{sect:smooth}. This topos contains a mirror
image of~$A$ which is a field, even if~$A$ is not.
\item $\mathrm{Bohr}(A)$, the \emph{Bohr topos} associated to a noncommutative
C\textsuperscript{*}\kern-.1ex-algebra~$A$. This topos contains a mirror image
of~$A$ which is commutative. In this sense, quantum mechanical systems (which are
described by noncommutative C\textsuperscript{*}\kern-.1ex-algebras) can be
regarded as classical mechanical systems (which are described by commutative
algebras)~\cite{butterfield-hamilton-isham:bohr,heunen-landsman-spitters:aqt}.
\item $\Set[\TT]$, the \emph{classifying topos} of a geometric theory~$\TT$.\footnote{A
geometric theory is a theory in many-sorted first-order logic whose axioms can be put as \emph{geometric
sequents}, sequents of the form~$\varphi \seq{\vec x} \psi$ where~$\varphi$
and~$\psi$ are geometric formulas (formulas built from equality and specified
relation symbols by the logical
connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$ and by arbitrary
set-indexed disjunctions~$\bigvee$).} This topos contains the
\emph{generic~$\TT$-model}. For instance, the classifying topos of the theory
of groups contains the \emph{generic group}. Arguably it is this group
which we implicitly refer to when we utter the phrase ``Let~$G$ be a group.''.
The generic group has has exactly those
properties which are shared by any group whatsoever.\footnote{More precisely,
this is only true for those properties which can be formulated as geometric
sequents. For arbitrary properties~$\varphi$, the statements~``the generic
group has property~$\varphi$'' and~``all groups have property~$\varphi$'' need
not be equivalent. This situation is explored
in~\cite{blechschmidt:nullstellensatz}.}
% FUTURE: Write just a little more on Set[T], that geom.impl. consequences spread.
% FUTURE: Give Set[T] its own section.
\item $T(\mathcal{L}_0)$, the \emph{free topos}. A statement is true in the free topos iff it is
intuitionistically provable. Lambek and Scott proposed that the free topos can
reconcile moderate platonism (because this topos has a certain universal
property which can be used to single it out among the plenitude of toposes),
moderate formalism (because it is constructed in a purely syntactic way) and
moderate logicism (because, as a topos, it supports an intuitionistic type
theory)~\cite{lambek:incompatible,couture-lambek:reflections}.
\end{enumerate}

@@

To obtain just a hint of how the true landscape looks like, we can study topos
theory from the inside of toposes; the resulting picture can look quite
different than the picture which emerges from within the standard topos.

@@

A basic observation in computability theory is that there are computational tasks
which are not solvable even for these idealized Turing machines. The paramount
example is the \emph{halting problem}: Given a Turing machine~$M$, determine
whether~$M$ terminates (comes to a stop after having carried out a finite
number of computational steps) or not.

@@

\begin{ex}\textbf{``Any function~$\RR \to \RR$ is continuous.''}\label{sect:eff-continuous}
In the standard topos, this statement is plainly false, with the sign and Heaviside functions
being prominent counterexamples. In the \effective topos, this statement is
true and independently due to
Kreisel--Lacombe--Shoenfield~\cite{kreisel-lacombe-shoenfield:cont} and
Ceĭtin~\cite{ceitin:cont}. A rigorous proof is not entirely
straightforward (a textbook reference
is~\cite[Theorem~9.2.1]{longley-normann:higher-order-computability}), but an intuitive
explanation is as follows.

@@

A function~$f : \RR \to \RR$ in the \effective topos is therefore given by a
Turing machine~$M$ which, when given the description of such a Turing machine~$X$ as
input, outputs the description of a similar such Turing machine~$Y$.
To compute a rational approximation~$Y(n)$, the machine~$Y$ may simulate~$X$
and can therefore determine arbitrarily many rational approximations~$X(m)$.
However, within in a finite amount of time, the machine~$Y$ can only learn finitely many
such approximations. Hence a function such as the sign function, for which
even rough rational approximations of~$\operatorname{sgn}(x)$ require infinite
precision in the input~$x$, do not exist in the \effective topos.
\end{ex}

@@

As a consequence, the input given to machine~$S_F$ is entirely uninformative
and~$S_F$ cannot make direct computational use of it. But its existence ensures that an
\emph{unbounded search} will not fail (and hence succeed, by an appeal to Markov's
principle on the meta level): The machine~$S_F$ can simulate~$F$ to
compute the values~$f(0), f(1), f(2), \ldots$ in turn, and stop
with output~$n$ as soon as it determines that some function value~$f(n)$ is zero.
\end{ex}

@@

  \caption{\label{table:eff} A (fragment of) the translation
  rules defining the meaning of statements internal to the \effective topos.}
\end{table}

@@

In the realizability topos constructed using infinite-time Turing machines, the full
law of excluded middle still fails, but some instances which are wrong in the
\effective topos do hold in this topos. For example, the instance ``any
function~$\NN \to \NN$ is the zero function or not'' does: Its external meaning
is that there is an infinite-time Turing machine~$M$ which, when given the
description of an infinite-time Turing machine~$F$ computing a function~$f :
\NN \to \NN$ as input, terminates (at some ordinal time step) with a correct
judgment of whether~$f$ is the zero function or not. Such a machine~$M$ indeed
exists: It simply has to simulate~$F$ on all inputs~$0,1,\ldots$ in order and
check whether one of the resulting function values is not zero. This
will require a transfinite amount of time (not least because simulating~$F$ on
just one input might require a transfinite amount of time), but as an
infinite-time Turing machine,~$M$ is capable of carrying out this procedure.

@@

A map~$\RR \to \NN$ in this topos is hence given by an infinite-time
Turing machine~$M$ which, when given the description of such an infinite-time
Turing machine~$X$ as input, outputs a certain natural number~$M(X)$. If~$X$
and~$X'$ represent the same real, then~$M(X)$ has to coincide with~$M(X')$.
This map~$\RR \to \NN$ is injective iff conversely,~$M(X) = M(X')$ implies
that~$X$ and~$X'$ represent the same real.

@@

The number~$M(X)$ computed by~$M$ depends on the input/output behaviour of~$X$,
the chosen order of infinite-time Turing machines, and on details of the
interleaving simulation and comparison procedure -- but it does not depend on
the implementation of~$X$ or on its specific choice of rational
approximations~$X(n)$. The search terminates since there is at least one
infinite-time Turing machine which represents the same real number as~$X$ does,
namely~$X$ itself.}

@@

Such a function is everywhere positive on~$X$ if and only if, from the internal point of
view~$\Sh(X)$, the number~$a$ is positive; it is everywhere zero if and only
if, internally, the number~$a$ is zero; and it is everywhere negative if and
only if, internally, the number~$a$ is negative.

@@

The weaker statement ``for any real number~$a$ it is \notnot the case that~$a < 0$
or~$a = 0$ or~$a > 0$'' does hold in~$\Sh(X)$, for this statement is an
intuitionistic tautology. Its meaning is that there exists a
dense open~$U$ such that~$U$ can be covered by opens on which~$a$ is either
everywhere negative, everywhere zero or everywhere positive. In the example
given in Figure~\ref{fig:trichotomy}(c), this open~$U$ could be taken as~$X$
with the unique zero of~$a$ removed.

@@

  \caption{\label{table:sheaf} A (fragment of) the translation rules defining
  the meaning of statements internal to~$\Sh(X)$, the topos of sheaves over a
  topological space~$X$.}
\end{table}

\begin{figure}
  \includegraphics[height=3cm]{trichotomy-1}
  \includegraphics[height=3cm]{trichotomy-2}
  \includegraphics[height=3cm]{trichotomy-3}
  \caption{\label{fig:trichotomy} Three examples of what the topos~$\Sh(X)$
  each believes to be a single real number, where the base space~$X$ is the
  unit interval. (a) A positive real number. (b) A negative real number. (c) A
  number which is neither negative nor zero nor positive. Externally speaking,
  there is no covering of the unit interval by opens on which the
  depicted function~$a$ is either everywhere negative, everywhere zero or everywhere
  positive.}
\end{figure}

@@

For reference purposes only, we include the precise definition of the Zariski topos.
\begin{defn}The \emph{Zariski topos of~$\RR$},~$\Zar(\RR)$, is a certain full subcategory of
the category of functors from finitely presented~$\RR$-algebras to sets, namely of
the Zariski sheaves. Such a functor is a \emph{Zariski sheaf} if and only if,
for any covering~$(A[f_i^{-1}])_i$ of any finitely presented~$\RR$-algebra~$A$
(this notion is defined in Table~\ref{table:zariski}), the diagram
\[ F(A) \longrightarrow \prod_i F(A[f_i^{-1}]) \xbigtoto{} \prod_{j,k}
F(A[(f_jf_k)^{-1}]) \]
is a coequalizer diagram. The object~$\RR^\sim$ of~$\Zar(\RR)$ is the tautologous functor~$A \mapsto A$.
\end{defn}

@@

    \bigskip
    \justify
    A \emph{covering} of an~$\RR$-algebra~$A$ is a family of~$A$-algebras of the
    form~$(A[f_i^{-1}])_i$ such that~$1 = f_1 + \cdots + f_n \in A$.
  \end{framed}
  \bigskip

  \caption{\label{table:zariski} A (fragment of) the Kripke--Joyal translation
  rules of the Zariski topos~$\Zar(\RR)$.}
\end{table}

@@

\subsection{On the importance of language} The verification of the field
property of~$\RR^\sim$ in Section~\ref{sect:the-zariski-topos} on
page~\pageref{page:field-property} demonstrates a basic feature of the
Kripke--Joyal translation rules: The translation~``$\E \models \varphi$'' of a
statement~$\varphi$ is usually quite complex, even if~$\varphi$ is reasonably
perspicuous.

@@

>>>>

The only caveat of Theorem~\ref{thm:reasoning} is that toposes generally only
support intuitionistic reasoning and not the full power of the ordinary
\emph{classical reasoning}. That is, within most toposes, the law of excluded
middle ($\varphi \vee \neg\varphi$) and the law of double negation elimination
($\neg\neg\varphi \Rightarrow \varphi$) are not available. It is intuitionistic
logic and not classical logic which is the common denominator of all toposes;
we cannot generally argue by contradiction in a topos.

@@

\begin{enumerate}
\item The \emph{trivial topos}. In the trivial topos, any statement whatsoever
is true. The trivial topos is not interesting on its own, but its existence
streamlines the theory and it can be an interesting question whether a given
topos coincides with the trivial topos.
\item $\Set$, the \emph{standard topos}. A statement is true in~$\Set$ iff it is true
in the ordinary mathematical sense.
\item $\Set_M$, the topos associated to any model~$(M,\in)$ of~\textsc{zf}.
\item $\Set^W$, the category of functors~$(W,\leq) \to \Set$ associated to
any Kripke model~$(W,\leq)$. A statement is true in this topos iff it is valid
with respect to the ordinary Kripke semantics of~$(W,\leq)$. This example shows
that the Kripke--Joyal semantics of toposes generalizes the more familiar
Kripke semantics.
\item $\Eff$, the \emph{\effective topos}. A statement is true in~$\Eff$ iff
if it has a \emph{computable witness} as detailed in
Section~\ref{sect:effective-topos}. In~$\Eff$, any function~$\NN \to \NN$ is
computable, any function~$\RR \to \RR$ is continuous and the countable axiom of
choice holds (even if it does not on the meta level).
\item $\Sh(X)$, the \emph{topos of sheaves} over any space~$X$. A statement is true
in~$\Sh(X)$ iff it holds \emph{locally on~$X$}, as detailed in
Section~\ref{sect:sheaf-toposes}. For most choices of~$X$, the
axiom of choice and the intermediate value theorem fail in~$\Sh(X)$, and this
failure is for geometric reasons.
\item $\Zar(A)$, the \emph{Zariski topos} of a ring~$A$ presented in
Section~\ref{sect:smooth}. This topos contains a mirror
image of~$A$ which is a field, even if~$A$ is not.
\item $\mathrm{Bohr}(A)$, the \emph{Bohr topos} associated to a noncommutative
C\textsuperscript{*}\kern-.1ex-algebra~$A$. This topos contains a mirror image
of~$A$ which is commutative. In this sense, quantum mechanical systems (which are
described by noncommutative C\textsuperscript{*}\kern-.1ex-algebras) can be
regarded as classical mechanical systems (which are described by commutative
algebras)~\cite{butterfield-hamilton-isham:bohr,heunen-landsman-spitters:aqt}.
\item $\Set[\TT]$, the \emph{classifying topos} of a geometric theory~$\TT$.\footnote{A
geometric theory is a theory in many-sorted first-order logic whose axioms can be put as \emph{geometric
sequents}, sequents of the form~$\varphi \seq{\vec x} \psi$ where~$\varphi$
and~$\psi$ are geometric formulas (formulas built from equality and specified
relation symbols by the logical
connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$ and by arbitrary
set-indexed disjunctions~$\bigvee$).} This topos contains the
\emph{generic~$\TT$-model}. For instance, the classifying topos of the theory
of groups contains the \emph{generic group}. Arguably it is this group
which we implicitly refer to when we utter the phrase ``Let~$G$ be a group.''.
The generic group has exactly those
properties which are shared by any group whatsoever.\footnote{More precisely,
this is only true for those properties which can be formulated as geometric
sequents. For arbitrary properties~$\varphi$, the statements~``the generic
group has property~$\varphi$'' and~``all groups have property~$\varphi$'' need
not be equivalent. This situation is explored
in~\cite{blechschmidt:nullstellensatz}.}
% FUTURE: Write just a little more on Set[T], that geom.impl. consequences spread.
% FUTURE: Give Set[T] its own section.
\item $T(\mathcal{L}_0)$, the \emph{free topos}. A statement is true in the free topos iff it is
intuitionistically provable. Lambek and Scott proposed that the free topos can
reconcile moderate platonism (because this topos has a certain universal
property which can be used to single it out among the plenitude of toposes),
moderate formalism (because it is constructed in a purely syntactic way) and
moderate logicism (because, as a topos, it supports an intuitionistic type
theory)~\cite{lambek:incompatible,couture-lambek:reflections}.
\end{enumerate}

@@

To obtain just a hint of how the true landscape looks like, we can study topos
theory from the inside of toposes; the resulting picture can look quite
different from the picture which emerges from within the standard topos.

@@

A basic observation in computability theory is that there are computational tasks
which are not solvable even for these idealized Turing machines. The paramount
example is the \emph{halting problem}: Given a Turing machine~$M$, determine
whether~$M$ terminates (comes to a stop after having carried out finitely
many computational steps) or not.

@@

\begin{ex}\textbf{``Any function~$\RR \to \RR$ is continuous.''}\label{sect:eff-continuous}
In the standard topos, this statement is plainly false, with the sign and Heaviside step functions
being prominent counterexamples. In the \effective topos, this statement is
true and independently due to
Kreisel--Lacombe--Shoenfield~\cite{kreisel-lacombe-shoenfield:cont} and
Ceĭtin~\cite{ceitin:cont}. A rigorous proof is not entirely
straightforward (a textbook reference
is~\cite[Theorem~9.2.1]{longley-normann:higher-order-computability}), but an intuitive
explanation is as follows.

@@

A function~$f : \RR \to \RR$ in the \effective topos is therefore given by a
Turing machine~$M$ which, when given the description of such a Turing machine~$X$ as
input, outputs the description of a similar such Turing machine~$Y$.
To compute a rational approximation~$Y(n)$, the machine~$Y$ may simulate~$X$
and can therefore determine arbitrarily many rational approximations~$X(m)$.
However, within a finite amount of time, the machine~$Y$ can only learn finitely many
such approximations. Hence a function such as the sign function, for which
even rough rational approximations of~$\operatorname{sgn}(x)$ require infinite
precision in the input~$x$, does not exist in the \effective topos.
\end{ex}

@@

As a consequence, the input given to the machine~$S_F$ is entirely uninformative
and~$S_F$ cannot make direct computational use of it. But its existence ensures that an
\emph{unbounded search} will not fail (and hence succeed, by an appeal to Markov's
principle on the meta level): The machine~$S_F$ can simulate~$F$ to
compute the values~$f(0), f(1), f(2), \ldots$ in turn, and stop
with output~$n$ as soon as it determines that some function value~$f(n)$ is zero.
\end{ex}

@@

  \caption{\label{table:eff}A (fragment of) the translation
  rules defining the meaning of statements internal to the \effective topos.}
\end{table}

@@

In the realizability topos constructed using infinite-time Turing machines, the full
law of excluded middle still fails, but some instances which are wrong in the
\effective topos do hold in this topos. For example, the instance ``any
function~$\NN \to \NN$ is the zero function or not'' does: Its external meaning
is that there is an infinite-time Turing machine~$M$ which, when given the
description of an infinite-time Turing machine~$F$ computing a function~$f :
\NN \to \NN$ as input, terminates (at some ordinal time step) with a correct
judgment of whether~$f$ is the zero function or not. Such a machine~$M$ indeed
exists: It simply has to simulate~$F$ on all inputs~$0,1,\ldots$ and
check whether one of the resulting function values is not zero. This search
will require a transfinite amount of time (not least because simulating~$F$ on
just one input might require a transfinite amount of time), but infinite-time
Turing machines are capable of carrying out this procedure.

@@

A map~$\RR \to \NN$ in this topos is hence given by an infinite-time
Turing machine~$M$ which, when given the description of such an infinite-time
Turing machine~$X$ as input, outputs a certain natural number~$M(X)$. If~$X$
and~$X'$ represent the same real, then~$M(X)$ has to coincide with~$M(X')$.
This map~$\RR \to \NN$ is injective iff conversely~$M(X) = M(X')$ implies
that~$X$ and~$X'$ represent the same real.

@@

The number~$M(X)$ computed by~$M$ depends on the input/output behaviour of~$X$,
the chosen order of infinite-time Turing machines, and on details of the
interleaving simulation and the comparison procedure -- but it does not depend on
the implementation of~$X$ or on its specific choice of rational
approximations~$X(n)$. The search terminates since there is at least one
infinite-time Turing machine which represents the same real number as~$X$ does,
namely~$X$ itself.}

@@

Such a function is everywhere positive on~$X$ if and only if, from the internal point of
view of~$\Sh(X)$, the number~$a$ is positive; it is everywhere zero if and only
if, internally, the number~$a$ is zero; and it is everywhere negative if and
only if, internally, the number~$a$ is negative.

@@

The weaker statement ``for any real number~$a$ it is \notnot the case that~$a < 0$
or~$a = 0$ or~$a > 0$'' does hold in~$\Sh(X)$, for this statement is a
theorem of intuitionistic calculus. Its meaning is that there exists a
dense open~$U$ such that~$U$ can be covered by opens on which~$a$ is either
everywhere negative, everywhere zero or everywhere positive. In the example
given in Figure~\ref{fig:trichotomy}(c), this open~$U$ could be taken as~$X$
with the unique zero of~$a$ removed.

@@

  \caption{\label{table:sheaf}A (fragment of) the translation rules defining
  the meaning of statements internal to~$\Sh(X)$, the topos of sheaves over a
  topological space~$X$.}
\end{table}

\begin{figure}
  \includegraphics[height=3cm]{trichotomy-1}
  \includegraphics[height=3cm]{trichotomy-2}
  \includegraphics[height=3cm]{trichotomy-3}
  \caption{\label{fig:trichotomy}Three examples of what the topos~$\Sh(X)$
  believes to be a single real number, where the base space~$X$ is the
  unit interval. (a) A positive real number. (b) A negative real number. (c)~A~number
  which is neither negative nor zero nor positive. Externally speaking,
  there is no covering of the unit interval by opens on which the
  depicted function~$a$ is either everywhere negative, everywhere zero or everywhere
  positive.}
\end{figure}

@@

For reference purposes only, we include the precise definition of the Zariski topos.
\begin{defn}The \emph{Zariski topos of~$\RR$},~$\Zar(\RR)$, is a certain full subcategory of
the category of functors from finitely presented~$\RR$-algebras to sets, namely of
the Zariski sheaves. Such a functor is a \emph{Zariski sheaf} if and only if,
for any covering~$(A[f_i^{-1}])_i$ of any finitely presented~$\RR$-algebra~$A$
(this notion is defined in Table~\ref{table:zariski}), the diagram
\[ F(A) \longrightarrow \prod_i F(A[f_i^{-1}]) \xbigtoto{} \prod_{j,k}
F(A[(f_jf_k)^{-1}]) \]
is an equalizer diagram. The object~$\RR^\sim$ of~$\Zar(\RR)$ is the tautologous functor~$A \mapsto A$.
\end{defn}

@@

    \bigskip
    \justify
    A \emph{covering} of an~$\RR$-algebra~$A$ is a finite family of~$A$-algebras of the
    form~$(A[f_i^{-1}])_{i=1,\ldots,n}$ such that~$1 = f_1 + \cdots + f_n \in A$.
  \end{framed}
  \bigskip

  \caption{\label{table:zariski}A (fragment of) the Kripke--Joyal translation
  rules of the Zariski topos~$\Zar(\RR)$.}
\end{table}

@@

\subsection{On the importance of language} The verification of the field
property of~$\RR^\sim$ in Section~\ref{sect:the-zariski-topos} on
page~\pageref{page:field-property} demonstrates a basic feature of the
Kripke--Joyal translation rules: The translation~``$\E \models \varphi$'' of a
statement~$\varphi$ is usually quite complex, even if~$\varphi$ is reasonably
transparent.

@@

XXX: Kleene-Vesley topos?
